---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

# Problem B

## Part (a)

```{r}
# Simulate the data for the medical screening experiment.
# Variables
# n :number of individuals, p: proportion of individuals with disease.
simulated_data <- function(n,p){
  state = rbinom(n =n, size=1, prob=p)
  diseased = sum(state) # number of diseased people
  normal = n - diseased # number of normal people
  protein_diseased = rgamma(diseased, shape = 2,scale = 1) # sample proteins 'diseased' number of times 
  protein_normal = rgamma(normal, shape = 2, scale = 0.5 ) # sample proteins 'normal' number of times 
  # make a dataframe
  protein_conc = c(protein_diseased,protein_normal)
  states = c(rep(1,diseased), rep(0,normal))
  df = data.frame(protein_conc, states)
  return(df)
} 

```

## Part (b)

```{r}
# The function takes the protein concetration as input and return likelihood ratio
LR <- function(protein_conc){
  lr = dgamma(protein_conc,scale = 0.5,shape = 2)/dgamma(protein_conc,scale = 1,shape = 2)
  return(lr)
}
```

## Part (c)

```{r}
data = simulated_data(1000000,0.2) # using 1 million data points
num_diseased = sum(data[,2]) # number of people diseased
lr = LR(data[,1]) # compute all likelihood ratios based on simulated protein concentrations
c_list = seq(1/10,10,length=1000)
tolerance = 0.2 # if the LR falls within the range [c - 0.2|c|, c + 0.2|c|] then, it's considered "near c"
# for each c, compute q_c
q_c_list = rep(0,length(c_list))
for (i in 1:length(c_list)){
  c = c_list[i]
  #print(c)
  near_c_range = c(c-tolerance*abs(c),c+tolerance*abs(c))
  #print(near_c_range)
  near_c = lr > near_c_range[1] & lr < near_c_range[2] # True if near c
  #print(near_c)
  near_c_and_diseased = sum(near_c[1:num_diseased])
  #print(near_c_and_diseased)
  # To avoid this computation producing NaN values, when near_c is 0 we conclude q_c = 0.
  if (sum(near_c)==0){
    q_c = 0
  }
  if (sum(near_c)!=0){
    q_c = near_c_and_diseased/sum(near_c)
  }
  #print(q_c)
  q_c_list[i] = q_c
}
```

Question: Among individuals with an LR "near" c, what proportion are truly diseased? 
Answer: The plot below shows that this proportion, $q_{D}(c)$, is a decreasing function of c by our construction of LR, the density for normal person divided by the density for diseased person at a particular protein concentration c. The proportion of truly diseased people can be obtained by reading y value of the curve at a particular x value.

Plot q_c against log10(c):

```{r}
plot(log10(c_list),q_c_list,type = 'l')
```

## Part (d) Plot

```{r}
# Theoretical q_c
q_c_theory <-function(c){
  q_c = 0.2/(0.8*c+0.2)
  return(q_c)
}

# Gather simulated data and theoretical values in one dataframe.

q_c_simul = q_c_list
q_c_thry = q_c_theory(c_list)
df_qc = data.frame(q_c_simul, q_c_thry)

# Plot simulated and theoretical values on the same panel
library(ggplot2)
ggplot() + 
  geom_line(data=df_qc, aes(y=q_c_simul, x=log(c_list), color = "darkred"), size=1) + 
  geom_line(data=df_qc, aes(y=q_c_thry, x=log(c_list), color = "steelblue"), size=1) +
  scale_colour_discrete(name = 'Experiments', labels = c("simulation", "theory"))
```

## Part (e)
```{r}
data2 = simulated_data(1000000,0.02) # Using 1 million datapoints
num_diseased = sum(data2[,2]) # number of people diseased
lr = LR(data2[,1]) # compute all likelihood ratios based on simulated protein concentrations
c_list = seq(1/10,10,length=1000)
tolerance = 0.2
q_c_list = rep(0,length(c_list))
for (i in 1:length(c_list)){
  c = c_list[i]
  near_c_range = c(c-tolerance*abs(c),c+tolerance*abs(c))
  near_c = lr > near_c_range[1] & lr < near_c_range[2] # True if near c
  near_c_and_diseased = sum(near_c[1:num_diseased])
  # To avoid this computation producing NaN values, when near_c is 0 we conclude q_c = 0.
  if (sum(near_c)==0){
    q_c = 0
  }
  if (sum(near_c)!=0){
    q_c = near_c_and_diseased/sum(near_c)
  }
  q_c_list[i] = q_c
}

q_c_theory2 <-function(c){
  q_c = 0.02/(0.98*c+0.02)
  return(q_c)
}

# Gather simulated data and theoretical values in one dataframe.

q_c_simul = q_c_list
q_c_thry = q_c_theory2(c_list)
df_qc = data.frame(q_c_simul, q_c_thry)

# Plot simulated and theoretical values on the same panel
library(ggplot2)
ggplot() + 
  geom_line(data=df_qc, aes(y=q_c_simul, x=log(c_list), color = "darkred"), size=1) + 
  geom_line(data=df_qc, aes(y=q_c_thry, x=log(c_list), color = "steelblue"), size=1) +
  scale_colour_discrete(name = 'Experiments', labels = c("simulation", "theory"))
```

# Problem C Part (b)

## Point estimate
```{r}
# Point estimates
shape = 12; scale = 1/9
## Mean
mean = shape*scale
## Median
median = qgamma(0.5,shape=12,scale=1/9)
print(paste("mean of the gamma distribution is ", mean))
print(paste("median of the gamma distribution is ", median))
```

## Interval estimate
```{r}
lower_bound = qgamma(0.05,shape=12,scale=1/9)
upper_bound = qgamma(0.95,shape=12,scale=1/9)
lower_bound
upper_bound
```

The 90% credible interval is: [`r lower_bound`,`r upper_bound`].

## Plot (just to check if this makes sense..)
```{r}
print(paste("The 90% credible interval is: "),lower_bound)
q = seq(0,10,length=100)
plot(q,dgamma(q,shape=12,scale=1/9),type = 'l')
```
