---
title: "Stat35450_HW5"
output:
  pdf_document: default
  html_document: default
---

## 2(b)

You can also embed plots, for example:

```{r}
m_ <- function(t){
  m = 0.1*(exp(0.01*t)-1)
  return(m)
}

i = 1
s = seq(from=0, to=1000, by=0.1)
for (t in s){
  i = i+1
  if (qpois(0.01,m_(t))==1000){
    print(m_(t))
    print(s[i])
  }
}
```

The above output implies that at $\lambda=1076$,  $P(N(t)\leq1000)\approx0.01$ and that the solution $t$ is some number around 928. To solve for $t$ exacctly we can use uniroot function to solve $m = 1076 =  0.1(e^{0.01t}-1)$ for $t$.

```{r}
uniroot(function(x) 0.1*(exp(0.01*x)-1)-1076, lower = -1, upper = 929, tol = 1e-9)

```

Thus, the answer is $t \approx 928.4$


## 2(d)

```{r}
m_ <- function(t){
  m = 0.05*(exp(0.01*t)-exp(-0.01*t))
  return(m)
}

i = 1
s = seq(from=0, to=1000, by=0.1)
for (t in s){
  i = i+1
  if (qpois(0.01,m_(t))==1000){
    print(m_(t))
    print(s[i])
  }
}
```

The above output implies that at $\lambda=1076$,  $P(N(t)\leq1000)\approx0.01$ and that the solution $t$ is some number around 998. To solve for $t$ exacctly we can use uniroot function to solve $m = 1076 =  0.05(e^{0.01t}-e^{-0.01t})$ for $t$.

```{r}
uniroot(function(x) 0.05*(exp(0.01*x)-exp(-0.01*x))-1076, lower = 997, upper = 999, tol = 1e-9)
```


Thus, the solution $t\approx997.7$


## 3(b)

For conveneience I let $\{0,1,2,3\}$ refers to ${A,C,T,G}$.We first define a function Pij_t that takes $i,j$ and $t$, where $i,j \in \{0,1,2,3\}$.

```{r}
Pij_t <- function(i,j,t){
  mu = 0.2*10^-8
  ft = (1 - exp(-mu*t))/4
  if (i==j){
    p_ij = 1 - ft
  }
  else if(i!=j){
    p_ij = ft
  }
  return(p_ij)
}

```

Now we can compute $P(AGG)$:

```{r}
sum = 0
for (i in c(0,1,2,3)){
  sum = sum + (Pij_t(i,0,6*10^6)^2)*(Pij_t(3,i,10^6))
  #print(sum)
}
P_AGG = Pij_t(3,3,7*10^6)*sum
P_AGG
log(P_AGG)
```

Thus the desired log-likelihood is `r log(P_AGG)`.

## 3(c)

```{r}
sum = 0
for (i in c(0,1,2,3)){
  sum = sum + Pij_t(0,i,10^6)*Pij_t(i,0,6*10^6)*Pij_t(i,3, 6*10^6)
  #print(sum)
}
P_AGA = Pij_t(0,0,7*10^6)*sum
P_AGA
log(P_AGA)
```


## 3(d)

Which tree is favored? Based on the above outputs, P_AGA model (from (c)) is favored over P_AGG model because the log of P_AGA is less negative than that of P_AGG. The difference in log-likelihood unit is:

```{r}
log(P_AGA) - log(P_AGG)
```
This results contradicts our knowledge that Chimpanzees are closer relatives to humans than Gorillas.Therefore, there needs to be more data collection to see if that changes the result in a way that's consistent with our knowledge.
